Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-22 report/MSc_Report.tex
.

Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty. Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices. A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals. Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background. We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction. All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.

This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence. Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required. The following sections outline our contributions, the research questions we address, and the scope of the work.

\section{Contributions and research questions}

We make three contributions within a single, notation-coherent framework.
\begin{enumerate}
  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
\end{enumerate}

Two questions motivate the empirical study.
\begin{enumerate}
  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines? We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)? We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
\end{enumerate}

\section{Scope and positioning}

We focus on triangular transport maps, transformation random forests, and copulas. Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions. Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used. Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions. Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.

\section{Takeaways and roadmap}

A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail. Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests. Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results. Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.

@@

We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$). All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated. This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.

\section{Change of Variables in Standardized Coordinates}

The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
\begin{equation}
  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
  \label{eq:pullback}
\end{equation}
Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate. Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$. The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions. This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.

It is useful to record the complementary pushforward statement. If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$. The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion. In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.

\section{Triangular Structure and Its Consequences}

A triangular map decomposes component-wise,
\begin{equation}
  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
  \label{eq:triangular-map}
\end{equation}
and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
\begin{equation}
  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
  \qquad
  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
  \label{eq:triangular-det}
\end{equation}

Three consequences matter in practice. First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$. Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}. Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components. Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.

Existence and anisotropy deserve a brief remark. For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}. Different orderings induce different maps. While all are valid, their sparsity and approximation difficulty may vary with order. When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling. In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.

Finally, we keep terminology consistent. A log density is $\log \hat\pi(\cdot)$ at a point. A (test) log likelihood is the dataset average of log densities. Its negative is the negative log likelihood (NLL) in nats; lower is better. These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.

\section{Separable and Cross-Term Triangular Transports}

A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.

\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
  \qquad
  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
  \label{eq:separable-component}
\end{equation}
so the log Jacobian depends only on $u_k$:
\begin{equation}
  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
  \label{eq:separable-jacobian}
\end{equation}
The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it. Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$. Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.

\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
\begin{equation}
  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
  \qquad
  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
  \label{eq:cross-component}
\end{equation}
so the log Jacobian, and thus the shape of the conditional, depends on the context. This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.

Two small examples make the contrast concrete. Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$. A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$. A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$. As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others. Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies. Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.

\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable. Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.

\section{Transformation Forests as Triangular Transport}

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}. A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}. This induces a triangular transport component via the probability integral transform,
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
  \label{eq:trtf-map-theory}
\end{equation}

>>>>

% Checklist: integrate unified transport framing, enforce US spelling, add dynamic refs.
Multivariate density estimation supports likelihood-based modeling, anomaly detection, simulation, and decision making under uncertainty. Tabular datasets often combine moderate to high dimension with context-dependent conditional structure. Conditional variance can change with predictors, and conditional skewness or modality can depend on earlier variables. A transport perspective addresses these challenges by coupling the target to a simple reference through an invertible map. This perspective enables exact likelihoods, transparent conditionals, and efficient sampling through a shared computational backbone \citep{rosenblatt1952remarks,knothe1957contributions}. This thesis compares three estimator families inside a single evaluation frame with matched objectives and diagnostics. The frame evaluates lower-triangular transport maps, Transformation Random Forests interpreted via probability integral transforms, and copulas that decouple marginals from dependence \citep{hothorn2017transformation,hothorn2018conditional,sklar1959fonctions}.
\medskip

We work in standardized coordinates to unify Jacobians, gradients, and reporting conventions. Let $x \in \mathbb{R}^K$ denote features on the original scale, and define
\begin{equation}
  u = T_{\text{std}}(x) = (x - \mu) \oslash \sigma,
  \qquad \sigma_k > 0,\label{eq:ch1-standardization}
\end{equation}
which standardizes each feature using training statistics only and fixes the coordinates where derivatives and Jacobians are evaluated. Let $S: u \mapsto z$ be a monotone lower-triangular map, and let $\eta$ denote the standard normal density. The change-of-variables identity gives
\begin{equation}
  \log \pi_U(u) = \log \eta\big(S(u)\big) + \log \big|\det \nabla_u S(u)\big|.
  \label{eq:ch1-pullback}
\end{equation}
Equation~\eqref{eq:ch1-pullback} splits the log density into a reference-fit term and a Jacobian term, isolating modeling capacity from the exact volume correction contributed by the map. For triangular $S$, the determinant factorizes as
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k} S_k(u_{1:k}).
  \label{eq:ch1-triangular-det}
\end{equation}
Equation~\eqref{eq:ch1-triangular-det} reduces the determinant to a sum of one-dimensional log derivatives, yielding linear per-sample complexity in $K$ and improving numerical stability. Reported log densities on the original scale apply the diagonal affine correction
\begin{equation}
  \log \pi_X(x) = \log \pi_U\!\big(T_{\text{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k.
  \label{eq:ch1-affine-correction}
\end{equation}
Equation~\eqref{eq:ch1-affine-correction} subtracts a constant offset determined by the training scales and ensures comparability across models that share standardized coordinates.
\medskip

We study separable triangular components for clarity, efficiency, and interpretability. Component $k$ decomposes into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),\qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k),
  \label{eq:ch1-separable}
\end{equation}
which stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.

% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
\section{Thesis and Problem Statement}\label{sec:ch1-problem}
This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame. We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests with axis-parallel splits (TRTF), and copula baselines.
\medskip

A transport perspective couples standardized data to a Gaussian reference through a monotone lower-triangular map. This structure yields exact likelihoods, transparent conditionals, exact inversion by back substitution, and linear per-sample evaluation. The Rosenblatt and Knothe rearrangements justify the triangular coupling for any variable order \citep{rosenblatt1952remarks,knothe1957contributions}.
\medskip

We standardize features using training statistics only. Equation~\eqref{eq:ch1-standardization} defines the standardized coordinates $u$ used for evaluation. All derivatives and Jacobians are computed in $u$. The diagonal affine correction in Equation~\eqref{eq:ch1-affine-correction} reports log densities on the original scale $x$. This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets. All negative log-likelihoods are reported in nats.
\medskip

We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\phi$ and $\Phi$. Triangular transport maps are abbreviated TTM, with the separable variant labeled TTM-Sep. Transformation Random Forests are abbreviated TRTF, and the axis-parallel implementation is labeled TRTF. Copulas decouple marginals from dependence and serve as interpretable baselines.
\medskip

Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}. The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant. TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor. Copulas preserve explicit marginals and model dependence on the unit hypercube.

We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute. The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost. Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.

Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes the pipeline by showing standardization $u = T_{\text{std}}(x)$, the triangular transport branch containing TTM-Sep and TRTF, and the copula branch. Both branches feed the reported outputs, namely log density, conditionals, sampling, calibration, and compute, under the shared frame.

The central problem is to determine when separability is appropriate for tabular data. We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention. Ordering effects, conditional calibration, and computational trade-offs address this question.
\medskip

On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MINIBOONE benchmark it improves on Gaussian references but trails published flow baselines. Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.

\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions. Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.

% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
\section{The Transport Frame on One Page}\label{sec:ch1-frame}
This section fixes the evaluation frame used across triangular transport maps, Transformation Random Forests, and copulas. It establishes notation, equations, and reporting conventions for the remainder of the thesis. The frame places all derivatives and Jacobians in standardized coordinates, applies a single diagonal affine correction on the original scale, and yields linear per-sample evaluation through a lower-triangular map.

Train-only standardization defines the evaluation coordinates via Equation~\eqref{eq:ch1-standardization}. The transformation uses training means and scales once and prevents information leakage into evaluation. All gradients and Jacobians act on $u$, which keeps objectives comparable across estimators and datasets. We convert to the original scale only when reporting.
\medskip

Let $S:u\mapsto z$ be a monotone lower-triangular map, and let $\eta(z)$ denote the $K$-variate standard normal density. The pullback identity in Equation~\eqref{eq:ch1-pullback} evaluates the reference at $S(u)$ and applies the exact volume correction. The split isolates model fit from geometry: the first term measures closeness to the Gaussian reference, and the second term contributes the exact Jacobian adjustment implied by the map.
\medskip

Lower-triangular structure with strictly positive diagonal partial derivatives factorizes the Jacobian as in Equation~\eqref{eq:ch1-triangular-det}. The factorization yields $\mathcal{O}(K)$ per-sample time for likelihood evaluation and improves numerical stability. It also guarantees global invertibility by back substitution consistent with the Rosenblatt and Knothe rearrangements.
\medskip

Reported log densities on the original scale apply only the diagonal affine correction implied by standardization. Equation~\eqref{eq:ch1-affine-correction} provides that adjustment. This convention keeps training and evaluation in $u$-space and converts to $x$-space at report time. All negative log-likelihoods are reported in nats, which fixes units across tables and figures.
\medskip

For clarity and efficiency we use separable triangular components of the form in Equation~\eqref{eq:ch1-separable}. Separable structure shifts location through $g_k$ and fixes conditional shape through the univariate monotone $h_k$. The Jacobian contribution depends only on $u_k$, which simplifies inversion by back substitution and stabilizes the log-determinant accumulation. The constraint also clarifies limits because conditional variance, skewness, and modality do not vary with context under separability.

The core operations follow the same path for all estimators.
\paragraph{Core operations.}
\begin{enumerate}
  \item First, standardize input features with training statistics using Equation~\eqref{eq:ch1-standardization}.
  \item Then, learn a monotone lower-triangular map $S$ to the standard normal reference and exploit Equation~\eqref{eq:ch1-pullback}.
  \item Finally, evaluate likelihoods, conditionals, and samples through Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction}.
\end{enumerate}

Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline. The input $x$ enters standardization to produce $u$. The triangular branch houses TTM-Sep and TRTF, whereas the copula branch couples fitted marginals to a dependence density. Both branches feed the same reported quantities, namely log density, conditionals, sampling, calibration, and compute. The appendix placement preserves the full schematic while keeping Chapter~\ref{ch:intro} focused on the narrative.

Notation remains consistent. We write $\eta$ for the $K$-variate standard normal density, and $\phi$ and $\Phi$ for the univariate standard normal density and CDF. We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$. These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.

This one-page frame removes duplicated exposition from Chapter~\ref{ch:background}. It establishes where logs and Jacobians live and makes complexity, inversion, and units explicit before the comparisons that follow. Section~\ref{sec:ch1-problem} documented the motivation, and Section~\ref{sec:ch1-contributions} states the resulting contributions and research questions.

% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
\section{Contributions and Research Questions}\label{sec:ch1-contributions}
This section states the contributions of the thesis and formulates the research questions. It also maps both elements to the chapters and figures that deliver the evidence. The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow. All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.

The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines. The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map. It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators. Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.

The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting. We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data. The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation. Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.

The third contribution distills practical guidance from the unified frame and the benchmarks. We state operational choices that preserve comparability, highlight ordering sensitivity and separability limits, and summarize when copulas serve as informative baselines. Chapter~\ref{ch:conclusion} consolidates these points as actionable recommendations and records limitations that motivate richer parameterizations or alternative predictors.

Two questions drive the empirical study and bind the contributions to specific measurements. The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data. All estimators share the transport frame in this comparison. We answer by reporting average test negative log-likelihoods in nats, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost. Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.

The second question asks how closely our TRTF results on real benchmarks approach the published performance of modern normalizing flows under the standard preprocessing. We answer by placing our test log-likelihoods beside reported numbers from the literature. The gaps are interpreted through the separable Jacobian constraint and compute profiles. Section~\ref{sec:realdata} reports these comparisons, and Chapter~\ref{ch:conclusion} interprets their implications for model choice.

Scope and non-goals maintain focus and ensure reproducibility. We study separable triangular maps and TRTF with additive predictors and compute all derivatives and Jacobians in standardized coordinates. The map direction $S:u \to z$ remains fixed for evaluation and inversion. Copulas include Gaussian dependence and a low-dimensional nonparametric variant used strictly as a diagnostic baseline. We treat high-capacity flows as external references rather than primary models, and we do not evaluate non-additive TRTF variants or cross-term triangular maps in this chapter. Chapter~\ref{ch:background} records the formal assumptions and notation. Section~\ref{sec:datasets-preprocessing} details preprocessing, seeds, and reporting conventions that keep results comparable across datasets and estimators.

Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.

@@

\section{Transport Frame and Notation}\label{sec:transport-frame}

This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis. The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters. We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.

We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
\begin{equation}
  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
\end{equation}
where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division. In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.

The standardised density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$. Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$. The pullback identity then reads
\begin{equation}
  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
\end{equation}
which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant. Reporting log densities on the original scale requires only the diagonal affine correction implied by standardisation,
\begin{equation}
  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
\end{equation}
We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.

The transport is assumed to be lower triangular and componentwise monotone,
\begin{equation}
  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
\end{equation}
so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorises into a sum of one-dimensional log derivatives,
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
\end{equation}
The factorisation yields $\mathcal{O}(K)$ evaluation cost per sample, improves numerical stability, and guarantees global invertibility: strict monotonicity lets us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.

Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame. All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting. The remainder of this chapter adopts this frame. Section~\ref{sec:transport-separable} details the separable triangular parameterisation used for direct transports. Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform. Section~\ref{sec:transport-copula} places copulas in the same reporting convention.

\begin{table}[t]
  \centering
  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}. All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
  \label{tab:transport-notation}
  \begin{tabular}{ll}
    \hline
    Symbol & Meaning \\ 
    \hline
    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
    $T_{\mathrm{std}}$ & Standardisation map using training $(\mu,\sigma)$ \\ 
    $u = T_{\mathrm{std}}(x)$ & Standardised evaluation coordinates \\ 
    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
    $\eta(z)$ & $K$-variate standard normal density \\ 
    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
    $K$ & Dimension of the feature vector \\ 
    \hline
  \end{tabular}
\end{table}

\section{Separable Triangular Maps and Transformation Random Forests as Transport}\label{sec:transport-separable}

% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}. Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}. We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

The goal is to state a single likelihood for both constructions, clarify what separability permits, and identify failure modes that motivate richer parameterizations. We do not pursue non-additive TRTF predictors, cross-term triangular maps, or ordering heuristics in this section; Chapter~\ref{ch:dataanalysis} evaluates those choices empirically and Appendix~\ref{ch:appendix} documents routines and defaults.

We adopt the notation introduced in Section~\ref{sec:transport-frame}. Coordinates satisfy $u=T_{\mathrm{std}}(x)$, the reference density is $\eta(z)$, and the pullback identity \eqref{eq:transport-pullback} gives $\pi_U(u)=\eta(S(u))\,|\det\nabla_u S(u)|$. The map is lower-triangular with strictly positive diagonal partial derivatives, which yields the sum decomposition in Equation~\eqref{eq:transport-det}. These conventions keep derivatives in $u$-space and apply the affine correction \eqref{eq:transport-affine} only when reporting $\log \pi_X(x)$.

We restrict attention to separable triangular maps. Component $k$ decomposes into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k),\qquad \log \partial_{u_k}S_k(u_{1:k}) \;=\; \log h_k'(u_k),\label{eq:transport-separable}
\end{equation}
which fixes context effects in $g_k$ and reserves $h_k$ for the one-dimensional marginal shape. Intuitively, earlier coordinates translate the location, while the conditional shape along $u_k$ remains fixed across contexts. The Jacobian contribution depends only on $u_k$, which reduces per-sample evaluation cost and simplifies inversion.

Substituting the standard normal reference into \eqref{eq:transport-pullback} produces a separable objective,
\begin{equation}
  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big],\label{eq:transport-likelihood}
\end{equation}
where $\varphi$ denotes the univariate standard normal density. Equation~\eqref{eq:transport-likelihood} splits the log density into a reference fit and an exact volume correction. In plain language, the model evaluates how Gaussian each transformed coordinate appears, then corrects for the local stretch induced by $h_k$. The same decomposition produces linear per-sample time in $K$ and stable accumulation of log derivatives.

The negative log-likelihood per sample takes the quadratic-plus-barrier form
\begin{equation}
  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
\end{equation}
which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out. Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier. In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

Separable structure encodes clear modeling assumptions. Conditional variance, skewness, and modality do not change with the preceding coordinates once $g_k$ shifts location. Consequently, separable maps can underfit heteroskedastic or multimodal conditionals, which manifests as U-shaped or inverted-U probability integral transform (PIT) diagnostics. Variable ordering also matters for finite bases because triangular transports are anisotropic, even though a Knothe--Rosenblatt rearrangement exists for any ordering \citep{rosenblatt1952remarks,knothe1957contributions}. These caveats guide the robustness checks in Chapter~\ref{ch:dataanalysis}.

\subsection{Transformation Random Forests within the Transport Frame}\label{sec:transport-trtf}
